Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
removed all red and blue coloring

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order {\em  eager} language, a straightforward  extension of the
earlier  technique  is  not  possible  for  lazy  languages,  where  a
heap-allocated   object   may   be  an   unevaluated   expression   or
closure. Since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.   In addition, apart  from evaluated data,  we extend
extend liveness-based  collection also to closures---closures  with no
future use should also be  collected.  This involves tracking closures
to their  points of  creation at  runtime from  where we  obtain their
liveness.  This  requires significant  modification to the  earlier GC
scheme.

@@

\begin{figure*}[t!]
\renewcommand{\arraystretch}{1.1}
  \begin{tabular}{cc}
    \begin{minipage}{.40\textwidth}
%        \small
        \renewcommand{\arraystretch}{1.1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1.1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow
(\abigfunction\ \acdr)}$ \IN
          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow
(}$\cred{+}$\cred{\ \pa\ \acdr)}$
\IN
	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ \cred{(\LET\ \pc}\
$\cred{\leftarrow  (\CONS\ \pb\ \NIL)}$ \IN
          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
          $\cred{(\LET\ \pw}\  $\cred{\leftarrow  (\length\ \pc)}$ \IN
          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

      %\small
\begin{eqnarray*}
        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
              [\;\epath_1 \mapsto \lbrace \px.\clazy \sigma,
\xl.\clazy\sigma\rbrace ,~\epath_2 \mapsto \lbrace \cred{
\pv.\sigma}  \rbrace,\\
 &&          \;\;  \epath_3 \mapsto \lbrace \pz.\sigma,
                \py.\clazy\sigma,
\pu.\Lf{\length}{\mbox{1}}{\clazy\sigma},\\
&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\clazy\sigma} \cup
%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
\cred{\clazy\Lf{\length}{1}{\clazy\sigma}}\}\}
             \rbrack \\
            \cal{P_{\mbox{\tiny \length}}} &=&
                [\;\pi_1 \mapsto \lbrace \epath_1, \epath_2, \epath_3 \rbrace,  \
                 \pi_2 \mapsto \lbrace \epath_1, \epath_2, \epath_3
                  \rbrace, \\
&&\;\;  \pi_3 \mapsto \lbrace \epath_2
                  \rbrace ,  \pi_4 \mapsto \lbrace \epath_2
                  \rbrace, \pi_5 \mapsto \lbrace  \epath_3 \rbrace, \\
 &&  \;\;                        \pi_6 \mapsto \lbrace \epath_3 \rbrace,
\pi_7 \mapsto
                   \lbrace \epath_3\rbrace, \pi_8 \mapsto \lbrace \epath_3
\rbrace~\rbrack\\
              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
                [\;
                  \epath_4 \mapsto \lbrace \cred {\pw.\sigma,
                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},}\\
&& \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
\cred{\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},}
\cred{\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
                               ] \\
               \cal{P_{\mbox{\tiny \main}}} &=&
                  \lbrack~ \cred{\pi_9 \mapsto \lbrace \epath_4 \rbrace,  \
                    \pi_{10} \mapsto \lbrace \epath_4 \rbrace,  \
                   \pi_{11} \mapsto \lbrace \epath_4 \rbrace,} \\
                   && \;\; \cred{\pi_{12} \mapsto \lbrace \epath_4 \rbrace,
\pi_{13} \mapsto \lbrace \epath_4 \rbrace}~\rbrack
\end{eqnarray*}
    \end{minipage} \\
(a)&
(b)
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
%\vspace{0.5in}
\caption{(a) Example program and (b) its liveness
maps.}\label{fig:mot-example2}
\end{figure*}
%================================================================

@@

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
{\color{blue}We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  senstive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFA. These DFA are consulted by the garbage collector
  to improve  collection.

@@

  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
  based garbage collector using heuristics  such as limiting the depth
  of  DFA,   merging  nearly-equivalent   states  and   using  better
  representation  and algorithms  for automata  manipulation. We  also
  need  to   investigate  the  interaction  of   liveness  with  other
  collection schemes, such as incremental and generational collection.
  In  summary, we  need to  investigate  ways to  make liveness  based
  garbage collection attractive for practical collectors.  }

>>>>

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order {\em  eager} language, a straightforward  extension of the
earlier  technique  is  not  possible  for  lazy  languages,  where  a
heap-allocated   object   may   be  an   unevaluated   expression   or
closure. Since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.   In addition, apart  from evaluated data,  we extend
liveness-based collection  to closures---closures  with no  future use
should also  be collected.  This  involves tracking closures  to their
points of  creation at  runtime from where  we obtain  their liveness.
This requires significant modification to the earlier GC scheme.

@@

\begin{figure*}[t!]
\renewcommand{\arraystretch}{1.1}
  \begin{tabular}{cc}
    \begin{minipage}{.40\textwidth}
%        \small
      \renewcommand{\arraystretch}{1.1}{
        \begin{uprogram}
          \UNL{1} (\DEFINE\ (\abigfunction~\py)
          \UNL{2} $(Create\ a\ big\ closure\ and\ return)$ 
          \end{uprogram}}
        \renewcommand{\arraystretch}{1.1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1.1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
           \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow
(\abigfunction\ \acdr)$ \IN
          \UNL{3} \!\!$\pi_{10}\!\!:\, (\LET\  \pb\  \leftarrow ($+$\ \pa\ \acdr)$
\IN
	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ (\LET\ \pc\
$\leftarrow  (\CONS\ \pb\ \NIL)$ \IN
          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
          $(\LET\ \pw\  $\leftarrow  (\length\ \pc)$ \IN
          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

      %\small
\begin{eqnarray*}
        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
              [\;\epath_1 \mapsto \lbrace \px.\clazy \sigma,
\xl.\clazy\sigma\rbrace ,~\epath_2 \mapsto \lbrace 
\pv.\sigma  \rbrace,\\
 &&          \;\;  \epath_3 \mapsto \lbrace \pz.\sigma,
                \py.\clazy\sigma,
\pu.\Lf{\length}{\mbox{1}}{\clazy\sigma},\\
&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\clazy\sigma} \cup
%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
\clazy\Lf{\length}{1}{\clazy\sigma}\}\}
             \rbrack \\
            \cal{P_{\mbox{\tiny \length}}} &=&
                [\;\pi_1 \mapsto \lbrace \epath_1, \epath_2, \epath_3 \rbrace,  \
                 \pi_2 \mapsto \lbrace \epath_1, \epath_2, \epath_3
                  \rbrace, \\
&&\;\;  \pi_3 \mapsto \lbrace \epath_2
                  \rbrace ,  \pi_4 \mapsto \lbrace \epath_2
                  \rbrace, \pi_5 \mapsto \lbrace  \epath_3 \rbrace, \\
 &&  \;\;                        \pi_6 \mapsto \lbrace \epath_3 \rbrace,
\pi_7 \mapsto
                   \lbrace \epath_3\rbrace, \pi_8 \mapsto \lbrace \epath_3
\rbrace~\rbrack\\
              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
                [\;
                  \epath_4 \mapsto \lbrace \pw.\sigma,
                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},\\
&& \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},
\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma} \rbrace
                               ] \\
               \cal{P_{\mbox{\tiny \main}}} &=&
                  \lbrack~ \pi_9 \mapsto \lbrace \epath_4 \rbrace,  \
                    \pi_{10} \mapsto \lbrace \epath_4 \rbrace,  \
                   \pi_{11} \mapsto \lbrace \epath_4 \rbrace, \\
                   && \;\; \pi_{12} \mapsto \lbrace \epath_4 \rbrace,
\pi_{13} \mapsto \lbrace \epath_4 \rbrace~\rbrack
\end{eqnarray*}
    \end{minipage} \\
(a)&
(b)
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
%\vspace{0.5in}
\caption{(a) Example program and (b) its liveness
maps.}\label{fig:mot-example2}
\end{figure*}
%================================================================

@@

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  senstive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFA. These DFA are consulted by the garbage collector
  to improve  collection.

@@

  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
  based garbage collector using heuristics  such as limiting the depth
  of  DFA,   merging  nearly-equivalent   states  and   using  better
  representation  and algorithms  for automata  manipulation. We  also
  need  to   investigate  the  interaction  of   liveness  with  other
  collection schemes, such as incremental and generational collection.
  In  summary, we  need to  investigate  ways to  make liveness  based
  garbage collection attractive for practical collectors.
